---
title: "Survival on the RMS Titanic"
author: "Przemyslaw Biecek"
date: "`r Sys.Date()`"
output: rmarkdown::html_document
vignette: >
  %\VignetteIndexEntry{Survival on the RMS Titanic}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = FALSE,
  comment = "#>",
  warning = FALSE,
  message = FALSE,
  cache = TRUE
)
```

# Data for Titanic survival

Let's see an example for `DALEX` package for Titanic passengers.
First, let's see the data, we will find quite nice data from in the `titanic` package.

```{r}
library("titanic")
titanic <- titanic_train[,c("Survived", "Pclass", "Sex", "Age", "SibSp", "Parch", "Fare", "Embarked")]
titanic$Survived <- factor(titanic$Survived)
titanic$Sex <- factor(titanic$Sex)
titanic$Embarked <- factor(titanic$Embarked)
levels(titanic$Embarked)[1] = "N"
titanic <- na.omit(titanic)
head(titanic)
```

# Model for Titanic survival

Ok, not it's time to create a model. Let's use the Random Forest model.

```{r}
# prepare model
library("randomForest")
rf_model <- randomForest(Survived ~ .,  data = titanic)
rf_model
```

# Explainer for Titanic survival

The third step (it's optional but useful) is to create a `DALEX` explainer for random forest model.

```{r}
library("DALEX")
rf_explain <- explain(rf_model, data = titanic[,-1],
                      y = titanic$Survived == "1", label = "Random Forest v7")
```

# Variable importance plots

Use the `variable_importance()` explainer to present importance of particular features. Note that `type = "difference"` normalizes dropouts, and now they all start in 0.

```{r}
vi_rf <- variable_importance(rf_explain, type = "difference")
head(vi_rf)
plot(vi_rf)
```

# Variable effects

As we see the most important feature is `Sex`. Next three importnat features are `Pclass`, `Age` and `Fare`. Let's see the link between model response and these features.

Such univariate relation can be calculated with `variable_response()`.

## Age

Kids 5 years old and younger have much higher survival probability.

```{r}
vr_age  <- variable_response(rf_explain, variable =  "Age")
head(vr_age)
plot(vr_age, use_facets = TRUE)
```

## Pclass

Passangers in the first class have much higher survival probability.

```{r}
vr_pclass  <- variable_response(rf_explain, variable =  "Pclass")
plot(vr_pclass, use_facets = TRUE)
```

## Fare

Very cheap tickets are linked with lower chances.

```{r}
vr_fare  <- variable_response(rf_explain, variable =  "Fare")
plot(vr_fare, use_facets = TRUE)
```

## Embarked

Passangers that embarked from C have highest survival.

```{r}
vr_embarked  <- variable_response(rf_explain, variable =  "Embarked")
plot(vr_embarked)
```


# Instance level explanations

Let's see break down explanation for model predictions for 8 years old male from 1st class that embarked from port C.

```{r}
new_passanger <- data.frame(
  Pclass = 1,
  Sex = factor("male", levels = c("female", "male")),
  Age = 8,
  SibSp = 0,
  Parch = 0,
  Fare = 72,
  Embarked = factor("C", levels = c("N","C","Q","S"))
)

sp_rf <- single_prediction(rf_explain, new_passanger)
plot(sp_rf)
```

It looks like the most important feature for this passenger is `Age` and `Sex`. After all his odds for survival are higher than for the average passenger. Mainly because of the young age and despite of being a male.

# Session info

```{r}
sessionInfo()
```
