---
title: "Survival on the RMS Titanic"
author: "Przemyslaw Biecek"
date: "`r Sys.Date()`"
output: rmarkdown::html_document
vignette: >
  %\VignetteIndexEntry{Survival on the RMS Titanic}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = FALSE,
  comment = "#>",
  warning = FALSE,
  message = FALSE
)
```

# Data for Titanic survival

Let's see an example for `DALEX` package for classification models for the survival problem for Titanic dataset.
Here we are using a dataset `titanic` avaliable in teh `DALEX` package. Note that this data was copied from the `stablelearner` package.

```{r}
library("DALEX")
head(titanic)
```

# Model for Titanic survival

Ok, not it's time to create a model. Let's use the Random Forest model.

```{r}
# prepare model
library("randomForest")
titanic <- na.omit(titanic)
model_titanic_rf <- randomForest(survived == "yes" ~ gender + age + class + embarked +
                                   fare + sibsp + parch,  data = titanic)
model_titanic_rf
```

# Explainer for Titanic survival

The third step (it's optional but useful) is to create a `DALEX` explainer for random forest model.

```{r}
library("DALEX")
explain_titanic_rf <- explain(model_titanic_rf, 
                      data = titanic[,-9],
                      y = titanic$survived == "yes", 
                      label = "Random Forest v7",
                      colorize = FALSE)
```

# Variable importance plots

Use the `variable_importance()` explainer to present importance of particular features. Note that `type = "difference"` normalizes dropouts, and now they all start in 0.

```{r}
vi_rf <- variable_importance(explain_titanic_rf)
head(vi_rf)
plot(vi_rf)
```

# Variable effects

As we see the most important feature is `Sex`. Next three importnat features are `Pclass`, `Age` and `Fare`. Let's see the link between model response and these features.

Such univariate relation can be calculated with `variable_effect()`.

## Age

Kids 5 years old and younger have much higher survival probability.

```{r}
vr_age  <- variable_effect(explain_titanic_rf, variables =  "age")
head(vr_age)
plot(vr_age)
```

## Passanger class

Passangers in the first class have much higher survival probability.

```{r}
vr_class  <- variable_effect(explain_titanic_rf, variables =  "class")
plot(vr_class)
```

## Fare

Very cheap tickets are linked with lower chances.

```{r}
vr_fare  <- variable_effect(explain_titanic_rf, variables =  "fare")
plot(vr_fare)
```

## Embarked

Passangers that embarked from C have highest survival.

```{r}
vr_embarked  <- variable_effect(explain_titanic_rf, variables =  "embarked")
plot(vr_embarked)
```


# Instance level explanations

Let's see break down explanation for model predictions for 8 years old male from 1st class that embarked from port C.

```{r}
new_passanger <- data.frame(
  class = factor("1st", levels = c("1st", "2nd", "3rd", "deck crew", "engineering crew", "restaurant staff", "victualling crew")),
  gender = factor("male", levels = c("female", "male")),
  age = 8,
  sibsp = 0,
  parch = 0,
  fare = 72,
  embarked = factor("Southampton", levels = c("Belfast", "Cherbourg", "Queenstown", "Southampton"))
)

sp_rf <- variable_attribution(explain_titanic_rf, new_passanger)
plot(sp_rf)
```

It looks like the most important feature for this passenger is `age` and `sex`. After all his odds for survival are higher than for the average passenger. Mainly because of the young age and despite of being a male.

# More models

Let's train more models for survival.
 
## Logistic regression

```{r}
library("rms")
model_titanic_lmr <- lrm(survived == "yes" ~ class + gender + rcs(age) + sibsp +
                   parch + fare + embarked, titanic)
explain_titanic_lmr <- explain(model_titanic_lmr, data = titanic, 
                       y = titanic$survived == "yes", 
                       predict_function = function(m,x) predict(m, x, type="fitted"),
                       label = "Logistic regression")
```
 
## Generalized Boosted Models (GBM)

```{r}
library("gbm")
model_titanic_gbm <- gbm(survived == "yes" ~ class + gender + age + sibsp +
                     parch + fare + embarked, data = titanic, n.trees = 15000)
explain_titanic_gbm <- explain(model_titanic_gbm, data = titanic, 
                       y = titanic$survived == "yes", 
                       predict_function = function(m,x) predict(m, x, n.trees = 15000, type = "response"),
                       label = "Generalized Boosted Models",
                       colorize = FALSE)
```
 
## Support Vector Machines (SVM)

```{r}
library("e1071")
model_titanic_svm <- svm(survived == "yes" ~ class + gender + age + sibsp +
                     parch + fare + embarked, data = titanic, 
             type = "C-classification", probability = TRUE)
explain_titanic_svm <- explain(model_titanic_svm, data = titanic, 
                       y = titanic$survived == "yes", 
                       label = "Support Vector Machines",
                       colorize = FALSE)
```
 
## k-Nearest Neighbours (kNN)

```{r}
library("caret")
model_titanic_knn <- knn3(survived == "yes" ~ class + gender + age + sibsp +
                     parch + fare + embarked, data = titanic, k = 5)
explain_titanic_knn <- explain(model_titanic_knn, data = titanic, 
                       y = titanic$survived == "yes", 
                       predict_function = function(m,x) predict(m, x)[,2],
                       label = "k-Nearest Neighbours",
                       colorize = FALSE)
```

## Variable performance

```{r, fig.width=5, fig.height=7}
vi_rf <- variable_importance(explain_titanic_rf)
vi_lmr <- variable_importance(explain_titanic_lmr)
vi_gbm <- variable_importance(explain_titanic_gbm)
vi_svm <- variable_importance(explain_titanic_svm)
vi_knn <- variable_importance(explain_titanic_knn)

plot(vi_rf, vi_lmr, vi_gbm, vi_svm, vi_knn, bar_width = 4)
```

## Single variable

```{r, fig.width=8, fig.height=5}
vr_age_rf  <- variable_effect(explain_titanic_rf, variables =  "age")
vr_age_lmr  <- variable_effect(explain_titanic_lmr, variables =  "age")
vr_age_gbm  <- variable_effect(explain_titanic_gbm, variables =  "age")
vr_age_svm  <- variable_effect(explain_titanic_svm, variables =  "age")
vr_age_knn  <- variable_effect(explain_titanic_knn, variables =  "age")
plot(vr_age_rf, vr_age_lmr, vr_age_gbm, vr_age_svm, vr_age_knn)
```

# Instance level explanations

```{r, fig.width=6, fig.height=3}
sp_rf <- variable_attribution(explain_titanic_rf, new_passanger)
plot(sp_rf)
sp_lmr <- variable_attribution(explain_titanic_lmr, new_passanger)
plot(sp_lmr)
sp_gbm <- variable_attribution(explain_titanic_gbm, new_passanger)
plot(sp_gbm)
sp_svm <- variable_attribution(explain_titanic_svm, new_passanger)
plot(sp_svm)
sp_knn <- variable_attribution(explain_titanic_knn, new_passanger)
plot(sp_knn)
```

# Session info

```{r}
sessionInfo()
```
